---
title: "Group 7"
author: "Michael Gilbride, Teun Hooft, and Emily Koskinen"
output: html_document
date: "2026-03-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
load("~/Documents/EM&LM/Group Assignment/Group_7.RData")
df <- DF

library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(nlme)
library(lme4)
```

# Introduction

This study investigates the progression of motor symptoms in patients diagnosed with Parkinson's Disease (PD). The primary aim is to understand how smoking status influences the disease trajectory. In addition, the aim is to understand how this evolution is influenced by key patient characteristics such as sex, age at diagnosis.The study enrolled 500 patients who were diagnosed with PD between 1995 and 2011. Motor assessments were conducted at baseline and then annually for a total follow-up period of 7 years. The disease trajectory is assessed by motor symptoms measured using the Unified Parkinson's Disease Rating Scale (UPDRS) Part III: Motor Examination score. This score ranges from 0 to 108, where higher values indicate more severe motor impairment.

# Descriptive Analysis

```{r data_fix, echo=FALSE}
df$sex_num <- ifelse(df$sex == "male", 0, 1) # For plotting purposes
df$id <- as.numeric(as.character(df$id))

df_wide <- df %>% 
  pivot_wider(
    names_from = Visit,
    values_from = UPDRS,
    names_prefix = "UPDRS_"
  )
df_wide <- df_wide[, c(1:8, 12, 9, 13, 11, 10)] # Rearrange the columns into the correct order
head(df_wide)
```

We have the following distributions in our data. The sexes of the participants are roughly equal, with just slightly more females than males. Age ranges from 41.1 to 88 and follows no clear distribution. The majority of our participants are non-smokers, with former smokers making up about half the number of non-smokers and current smokers just over a third the amount of non-smokers. Finally, at baseline we have $\sim500$ measurements, which drops to the range 306-339 for each subsequent visit. This means our data is unbalanced, which indicates that a continuous AR1, exponential serial, or Gaussian serial correlation structure will be needed.

```{r sum_stats, echo=FALSE}
plot1 <- ggplot(data = df, aes(x = sex_num)) +
  geom_histogram(stat = "bin", binwidth = 0.5) +
  stat_bin(binwidth = 0.5,
           aes(label=ifelse(after_stat(count)==0,"",after_stat(count))),
           geom="text", vjust=-.5) +
  ylim(0, 1525) + scale_x_continuous(breaks = c(0,1)) +
  labs(x = "Sex", y = "Count")

plot2 <- ggplot(data = df, aes(x = Age)) +
  geom_histogram(stat = "bin", binwidth = 0.5) +
  ylim(0,85) +
  labs(y = "Count")

plot3 <- ggplot(df, aes(x = smoking_status)) +
  geom_bar() +
  geom_text(stat = "count",
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    vjust = -0.5) +
  ylim(0, 1600) +
  labs(x = "Smoking Status", y = "Count")

plot4 <- ggplot(df, aes(x = Visit)) +
  geom_bar() +
  geom_text(stat = "count",
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    vjust = -0.5) +
  ylim(0, 550) +
  labs(y = "Count")

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)
```

The mean UPDRS score per visit for each smoking subgroup (non-, former, and current smoker) appears non-linear. The standard error also increases for each visit.

```{r smoking_plot, echo=FALSE}
all.patients.summary <- df %>% 
  group_by(smoking_status, Visit) %>% # Split the participants into their smoking status and evaluate the UPDRS by Visit
  summarise(
    mean_UPDRS = mean(UPDRS, na.rm=TRUE),
    sd_UPDRS = sd(UPDRS, na.rm=TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

split_by_smoking <- facet_wrap(~ smoking_status, nrow = 1,  # Splits output into three graphs, one per smoking group
             labeller = labeller(smoking_status = c("0" = "Non-smoker",
                                                    "1" = "Former smoker",
                                                    "2" = "Current smoker")))

ggplot(all.patients.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21, fill = "white", size = 2) +
  geom_errorbar(aes(ymin = mean_UPDRS - se_UPDRS,
                     ymax = mean_UPDRS + se_UPDRS), 
                width = 0.5, color = "blue") +
  # geom_smooth(method = "loess", color = "red", linewidth = 0.2) + # How to interpret result here?
  split_by_smoking +
  labs(x = "Visit", y = "UPDRS Score")
```

This spread can be seen in the following spaghetti plots from a random sample of participants.

```{r spaghetti_plot}
random_ids <- df %>%
  distinct(smoking_status, id) %>%
  group_by(smoking_status) %>%
  slice_sample(n = 12) # Gets 12 of each smoking status randomly by their id

df_sample <- df %>%
  semi_join(random_ids, by = c("smoking_status", "id")) # Pulls the values from df

ggplot(df_sample, aes(x = Visit, y = UPDRS, group = id, colour = factor(id))) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap(~ smoking_status, nrow = 1) +
  split_by_smoking + 
  labs(x = "Visit", y = "UPDRS Score")
```

# Statistical Analysis

The plot of squared OLS residuals against visit indicates that the variance is not constant, which means that we have heteroscedasticity. Thus, we get further evidence backing up the spaghetti plots that the progression of Parkinson's across the patients in this study is not uniform.

```{r ols_plot}
lin.reg <- lm(UPDRS ~ Visit, data = df)
sq.ols.residuals <- resid(lin.reg)^2

df2 <- df
df2$sq_ols_residuals <- sq.ols.residuals
df2$fitted <- fitted(lin.reg)

ggplot(df2, aes(x = Visit, y = sq_ols_residuals)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Visit", y = "Squared OLS Residuals")
```

From the correlation matrix of the UPDRS Scores below, we get the expected result that UPDRS Scores are less correlated the further along the measurement was taken.

```{r cor}
#colSums(is.na(df_wide)) # Around 180 missing measurements per visit after baseline
df_wide_UPDRS <- subset(df_wide, select = -c(1,2,3,4,5))
cor(df_wide_UPDRS, use = "pairwise.complete.obs")
```
SLIDE 268 ONWARDS FOR HERE. 













# Appendix

## Model Selection

We started with a simple model without any interactions and focus on whether our model needed random intercepts only, or both random intercepts and random slopes.

COMMENT, re. correlation structure.
nlme just started working, God knows why. Cannot decide between CAR1 and AR1, neither anova nor residual plots gives any difference. 

Why the confusion? The visits themselves are equally spaced (which implies AR1), but the measurements are sometimes missing so the measurements aren't actually evenly spaced (which implies CAR1 if the model 'skips' over NAs). 

Will use AR1 as it's simpler so if it's all the same it'll save some computing time. Can change it with a single letter later so it's not too big a deal.


COMMENT re. models from R packages
I have included both nlme and lme4 below. If nlme continues working fine, I would prefer it, as I cannot specify correlation structure in lmer() (though this may not be necessary as controlling for random effects can already account for correlation) nor use anova() for hypothesis testing.

I'm aslo geting that the boundary is singular and that the model is refit with ML instead of REML. Likely comes from too few observations per subject (lots of missing measurements.)

Can't model heteroskedasticity directly with lme4? Does this explain why my nlme models have lower AIC/BIC?

Will trim the fat later. nlme is looking like it'll be better.


```{r comp_rand_lme4}
# Random intercepts only
m1 <- lmer(UPDRS ~ Visit + smoking_status + Age + sex + (1 | id), data = df)
# Random intercepts and random slopes
m2 <- lmer(UPDRS ~ Visit + smoking_status + Age + sex + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_rand_nlme}
# Random intercepts only
m1 <- lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = ~ 1 | id, # Visit | id
                  weights = varExp(form = ~ Visit), # defines heteroskedasticity in Visit within the group.
                  correlation = corAR1(form = ~ 1 | id), # visits are equally spaced, but measurements are sometimes missing: AR1 or CAR1?
                  method = "REML",
                  data = df)

# Random intercepts and random slopes
m2 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)), # Removes intercept-slope covariance (gets this model to work!)
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```
COMMENT, re. AR(1) and random slopes
Have seen it said that random slopes and AR(1) are alternative ways to model within-subject dependence. I test with anova() below.
```{r within_subject_dependence1}
# No AR(1), just random slope
m1 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = ~ Visit | id,
                  weights = varExp(form = ~ Visit),
                  method = "REML",
                  data = df)
# AR(1) and random slope with now intercept-slope covariance
m2 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)

anova(m1, m2)
```
```{r within_subject_dependence2}
# No random slope, just AR(1)
m1 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = ~ 1 | id,
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# AR(1) and random slope with now intercept-slope covariance
m2 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)

anova(m1, m2)
```
The model with both random slope (but no intercept-slope covariance) and AR(1) has the lowest AIC and BIC, so we keep it.



As these models aren't nested, we consider the information criteria. As the more complicated model has lower values for both, this is the one we choose.

We then checked the interactions.

```{r comp_int1}
# Random intercepts and random slopes
m1 <- lmer(UPDRS ~ Visit + smoking_status + Age + sex + (Visit | id), data = df)
# RI + RS + all interactions
m2 <- lmer(UPDRS ~ Visit*smoking_status*Age*sex + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme1}
# No interactions
m1 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```

Comparing the model with no interactions with the one with all interactions, we again see by the information criteria that the more complicated model is needed.

We then narrowed down the number of interactions.

```{r comp_int2}
# Remove interaction between visit and everything else
m1 <- lmer(UPDRS ~ Visit + smoking_status*Age*sex + (Visit | id), data = df)
# RI + RS + all interactions
m2 <- lmer(UPDRS ~ Visit*smoking_status*Age*sex + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme2}
# Remove interaction between visit and everything else
m1 <-lme(UPDRS ~ Visit + smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```


```{r comp_int3}
# Remove interaction between age and everything else
m1 <- lmer(UPDRS ~ Visit*smoking_status*sex + Age + (Visit | id), data = df)
# RI + RS + all interactions
m2 <- lmer(UPDRS ~ Visit*smoking_status*Age*sex + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme3}
# Remove interaction between age and everything else
m1 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```


```{r comp_int4}
# Remove interaction between age/sex and everything else
m1 <- lmer(UPDRS ~ Visit*smoking_status + sex + Age + (Visit | id), data = df)
# Remove interaction between age and everything else
m2 <- lmer(UPDRS ~ Visit*smoking_status*sex + Age + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme4}
# Remove interaction between age/sex and everything else
m1 <-lme(UPDRS ~ Visit*smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```


```{r comp_int5}
# Remove age/sex from interaction but have them interact and
m1 <- lmer(UPDRS ~ Visit*smoking_status + sex*Age + (Visit | id), data = df)
# Remove interaction between age and everything else
m2 <- lmer(UPDRS ~ Visit*smoking_status*sex + Age + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme5}
# Remove age/sex from interaction but have them interact and
m1 <-lme(UPDRS ~ Visit*smoking_status + sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```
```{r comp_int6}
# Remove age/smoking from interaction
m1 <- lmer(UPDRS ~ Visit*sex + smoking_status + Age + (Visit | id), data = df)
# Remove interaction between age and everything else
m2 <- lmer(UPDRS ~ Visit*smoking_status*sex + Age + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme6}
# Remove age/smoking from interaction
m1 <-lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```
```{r comp_int7}
# Remove age/smoking from interaction
m1 <- lmer(UPDRS ~ Visit*sex + smoking_status + Age + (Visit | id), data = df)
# Remove age/smoking from interaction but make them interact
m2 <- lmer(UPDRS ~ Visit*sex + smoking_status*Age + (Visit | id), data = df)
anova(m1, m2)
```
```{r comp_int_nlme7}
# Remove age/smoking from interaction
m1 <-lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove age/smoking from interaction but make them interact
m2 <-lme(UPDRS ~ Visit*sex + smoking_status*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
anova(m1, m2)
```

This led to our selecting the following as the true model;

$$ y_{ij} = \big( \beta_0 + \text{b}_{i0} \big) + \big(\beta_1 + \text{b}_{i1}\big)\text{Visit}_{ij} + \beta_2\,\text{sex}_i + \beta_3\,\text{smoking}_i + \beta_4\,\text{age}_i + \beta_5\,\text{Visit}_{ij}\times\text{sex}_i + \epsilon_{ij}$$
where $\beta$ are our model's regression parameters, $i = 1,2,\dots,2765$, $j = 0,1,\dots,7$, b$_{i0}$ denotes the random intercept of each patient, and b$_{i1}$ denotes the random slope of each patient's progression.


```{r true_model_nlme}
true.model <- lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
summary(true.model)
```
```{r plot_tm}
plot(true.model)
```

## Hypothesis Testing

We perform a t-test for each of our regression parameters (coefficients) in our model
```{r t_test}
summary(true.model)$tTable
```
COMMENT: Need to interpret the results

We test the coefficients in our model. We start with checking for different mean trajectories between the sexes. 
COMMENT: But why, since we already have our model. Some of these have different DFs and can't be compared, what does this mean? 

```{r hypo_sex_test}
anova(true.model, Terms = c("Visit", "Visit:sex"))
```

































