---
title: "Group 7"
date: "2026-03-25"
output: 
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-2cm} 
fontsize: 9pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
load("~/Documents/EM&LM/Group Assignment/Group_7.RData")
df <- DF

library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(nlme)
library(lme4)
```

# Introduction

This study investigates the progression of motor symptoms in patients diagnosed with Parkinson's Disease (PD). The primary aim is to understand how smoking status influences the disease trajectory. In addition, the aim is to understand how this evolution is influenced by key patient characteristics such as sex, age at diagnosis.The study enrolled 500 patients who were diagnosed with PD between 1995 and 2011. Motor assessments were conducted at baseline and then annually for a total follow-up period of 7 years. The disease trajectory is assessed by motor symptoms measured using the Unified Parkinson's Disease Rating Scale (UPDRS) Part III: Motor Examination score. This score ranges from 0 to 108, where higher values indicate more severe motor impairment.

# Descriptive Analysis

```{r data_fix, include=FALSE}
df$sex_num <- ifelse(df$sex == "male", 0, 1) # For plotting purposes
df$id <- as.numeric(as.character(df$id))

df_wide <- df %>% 
  pivot_wider(
    names_from = Visit,
    values_from = UPDRS,
    names_prefix = "UPDRS_"
  )
df_wide <- df_wide[, c(1:8, 12, 9, 13, 11, 10)] # Rearrange the columns into the correct order
head(df_wide)
```

We have the following distributions in our data. The sexes of the participants are roughly equal, with just slightly more females than males. Age ranges from 41.1 to 88 and follows no clear distribution. The majority of our participants are non-smokers, with former smokers making up about half the number of non-smokers and current smokers just over a third the amount of non-smokers. Finally, at baseline we have $\sim500$ measurements, which drops to the range 306-339 for each subsequent visit. This means our data is unbalanced, which indicates that a continuous AR1, exponential serial, or Gaussian serial correlation structure will be needed.

```{r sum_stats, out.width="75%"}
plot1 <- ggplot(data = df, aes(x = sex_num)) +
  geom_histogram(stat = "bin", binwidth = 0.5) +
  stat_bin(binwidth = 0.5,
           aes(label=ifelse(after_stat(count)==0,"",after_stat(count))),
           geom="text", vjust=-.5) +
  ylim(0, 1525) + scale_x_continuous(breaks = c(0,1)) +
  labs(x = "Sex", y = "Count")

plot2 <- ggplot(data = df, aes(x = Age)) +
  geom_histogram(stat = "bin", binwidth = 0.5) +
  ylim(0,85) +
  labs(y = "Count")

plot3 <- ggplot(df, aes(x = smoking_status)) +
  geom_bar() +
  geom_text(stat = "count",
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    vjust = -0.5) +
  ylim(0, 1600) +
  labs(x = "Smoking Status", y = "Count")

plot4 <- ggplot(df, aes(x = Visit)) +
  geom_bar() +
  geom_text(stat = "count",
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    vjust = -0.5) +
  ylim(0, 550) +
  labs(y = "Count")

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)
```

The mean UPDRS score per visit for each smoking subgroup (non-, former, and current smoker) appears non-linear. The standard error also increases for each visit.

```{r smoking_plot, out.width="75%"}
smokers.summary <- df %>% 
  group_by(smoking_status, Visit) %>% # Split the participants into their smoking status and evaluate the UPDRS by Visit
  summarise(.groups = "drop_last",
    mean_UPDRS = mean(UPDRS, na.rm=TRUE),
    sd_UPDRS = sd(UPDRS, na.rm=TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

split_by_smoking <- facet_wrap(~ smoking_status,  # Splits output into three graphs, one per smoking group
             labeller = labeller(smoking_status = c("0" = "Non-smoker", "1" = "Former smoker", "2" = "Current smoker")))

ggplot(smokers.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21, fill = "white", size = 2) +
  geom_errorbar(aes(ymin = mean_UPDRS - se_UPDRS,
                     ymax = mean_UPDRS + se_UPDRS), 
                width = 0.5, color = "blue") +
  # geom_smooth(method = "loess", color = "red", linewidth = 0.2) + # How to interpret result here?
  split_by_smoking +
  labs(x = "Visit", y = "UPDRS Score")
```

```{r sex_plot}
sex.summary <- df %>% 
  group_by(sex_num, Visit) %>% # Split the participants into their smoking status and evaluate the UPDRS by Visit
  summarise(.groups = "drop_last",
    mean_UPDRS = mean(UPDRS, na.rm=TRUE),
    sd_UPDRS = sd(UPDRS, na.rm=TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

split_by_sex <- facet_wrap(~ sex_num, labeller = labeller(sex_num = c("0" = "Male", "1" = "Female")))

ggplot(sex.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21, fill = "white", size = 2) +
  geom_errorbar(aes(ymin = mean_UPDRS - se_UPDRS,
                     ymax = mean_UPDRS + se_UPDRS), 
                width = 0.5, color = "blue") +
  split_by_sex +
  labs(x = "Visit", y = "UPDRS Score")
```


This spread can be seen in the following spaghetti plots from a random sample of participants.

```{r spaghetti_plots, out.width="75%"}
random_smokers <- df %>% distinct(smoking_status, id) %>% 
  group_by(smoking_status) %>% slice_sample(n = 12)
df_sample_smokers <- df %>%
  semi_join(random_smokers, by = c("smoking_status", "id"))

random_sex <- df %>% distinct(sex_num, id) %>% 
  group_by(sex_num) %>% slice_sample(n = 12)
df_sample_sex <- df %>%
  semi_join(random_sex, by = c("sex_num", "id"))


ggplot(df_sample_smokers, aes(x = Visit, y = UPDRS, group = id, colour = factor(id))) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap(~ smoking_status, nrow = 1) +
  split_by_smoking + 
  labs(x = "Visit", y = "UPDRS Score")

ggplot(df_sample_sex, aes(x = Visit, y = UPDRS, group = id, colour = factor(id))) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap(~ sex_num, nrow = 1) +
  split_by_sex + 
  labs(x = "Visit", y = "UPDRS Score")
```

# Statistical Analysis

The plot of squared OLS residuals against visit indicates that the variance is not constant, which means that we have heteroscedasticity. Thus, we get further evidence backing up the spaghetti plots that the progression of Parkinson's across the patients in this study is not uniform.

```{r ols_plot, out.width="75%"}
lin.reg <- lm(UPDRS ~ Visit, data = df)
sq.ols.residuals <- resid(lin.reg)^2

df2 <- df
df2$sq_ols_residuals <- sq.ols.residuals
df2$fitted <- fitted(lin.reg)

ggplot(df2, aes(x = Visit, y = sq_ols_residuals)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Visit", y = "Squared OLS Residuals")
```

From the correlation matrix of the available UPDRS Scores below, we get the expected result that UPDRS Scores are less correlated the further along the measurement was taken.
\footnotesize
```{r cor}
df_wide_UPDRS <- subset(df_wide, select = -c(1,2,3,4,5))
as.tibble(cor(df_wide_UPDRS, use = "pairwise.complete.obs"))
```
\normalsize
SLIDE 268 ONWARDS FOR HERE. 













# Appendix

## Model Selection

We started with a simple model without any interactions and focus on whether our model needed random intercepts only, or both random intercepts and random slopes.

Discrete AR(1) (denoted by AR1) was chosen over continuous AR(1) (CAR1) as the visits themselves are equally spaced (which implies AR1), even thought the measurements are sometimes missing so the actual measurements aren't actually evenly spaced.

The package *nlme* was chosen to generate our model because *lme4* doesn't model heteroskedasticity and doesn't allow for a specified correlation structure, the latter not necessarily being a problem conceptually but ended up not working well owing to many subjects having lots of missing measurements.

First, we check the within-subject dependence of the basic model by comparing random intercepts, random slopes, and the AR1 correlation structure, as the latter two are alternative ways to model these conditions.

\footnotesize
```{r within_subject_dependence1}
# AR(1) and random intercepts only
m1 <- lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = ~ 1 | id, # Visit | id
                  weights = varExp(form = ~ Visit), # defines heteroskedasticity in Visit within the group.
                  correlation = corAR1(form = ~ 1 | id), # visits are equally spaced, but measurements are sometimes missing: AR1 or CAR1?
                  method = "REML",
                  data = df)

# AR(1) and both random intercepts and random slopes
m2 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)), # Removes intercept-slope covariance (gets this model to work!)
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("AR1 & Random intercepts vs. AR1 & random intercepts + slopes")
anova(m1, m2)
```
```{r within_subject_dependence2}
# No AR(1), just random intercepts and random slope
m1 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = ~ Visit | id,
                  weights = varExp(form = ~ Visit),
                  method = "REML",
                  data = df)
# AR(1) and random slope with no intercept-slope covariance
m2 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Random intercepts + slopes vs. AR1 & random intercepts + slopes")
anova(m1, m2)
```
\normalsize
As these models aren't nested, we consider the information criteria. The model with both random intercepts and slope (but with no intercept-slope covariance) and AR(1) has the lowest AIC and BIC, so we keep it.

We then checked for interactions between the terms.
\footnotesize
```{r comp_int_nlme1}
# No interactions
m1 <-lme(UPDRS ~ Visit + smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("No interactions vs. Interaction")
anova(m1, m2)
```
\normalsize

Comparing the model with no interactions with the one with all interactions, we again see by the information criteria that the more complicated model is needed. We then narrowed down the number of interactions.
\footnotesize
```{r comp_int_nlme2}
# Remove interaction between visit and everything else
m1 <-lme(UPDRS ~ Visit + smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Visit non-interacting vs. All interaction")
anova(m1, m2)
print("Chosen: All interactions")
```
```{r comp_int_nlme3}
# Remove interaction between age and everything else
m1 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# All interactions
m2 <-lme(UPDRS ~ Visit*smoking_status*sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Age non-interacting vs. All interactions")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
```{r comp_int_nlme4}
# Remove interaction between age/sex and everything else
m1 <-lme(UPDRS ~ Visit*smoking_status + sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Age & sex non-interacting vs. Age non-interacting")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
```{r comp_int_nlme5}
# Remove age/sex from interaction but have them interact and
m1 <-lme(UPDRS ~ Visit*smoking_status + sex*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Age & sex only interacting with each other vs. Age non-interacting")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
```{r comp_int_nlme6}
# Remove age/smoking from interaction
m1 <-lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove interaction between age and everything else
m2 <-lme(UPDRS ~ Visit*smoking_status*sex + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Age & smoking status non-interacting vs. Age non-interacting")
anova(m1, m2)
print("Chosen: Age & smoking status non-interacting")
```
```{r comp_int_nlme7}
# Remove age/smoking from interaction
m1 <-lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
# Remove age/smoking from interaction but make them interact
m2 <-lme(UPDRS ~ Visit*sex + smoking_status*Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
print("Age & smoking non-interacting vs. Age & smoking only interacting with each other")
anova(m1, m2)
print("Chosen: Age & smoking status non-interacting")
```
\normalsize

This led to our selecting the following as the true model;

$$ y_{ij} = \big( \beta_0 + \text{b}_{i0} \big) + \big(\beta_1 + \text{b}_{i1}\big)\text{Visit}_{ij} + \beta_2\,\text{sex}_i + \beta_3\,\text{smoking}_i + \beta_4\,\text{age}_i + \beta_5\,\text{Visit}_{ij}\times\text{sex}_i + \epsilon_{ij}$$
where $\beta$ are our model's regression parameters, $i = 1,2,\dots,2765$ denotes the number of participants, $j = 0,1,\dots,7$ denotes the number of visits, b$_{i0}$ denotes the random intercept of each patient, and b$_{i1}$ denotes the random slope of each patient's progression.

\footnotesize
```{r true_model_nlme, echo=TRUE, out.width="75%"}
true.model <- lme(UPDRS ~ Visit*sex + smoking_status + Age, 
                  random = list(id = pdDiag(~ Visit)),
                  weights = varExp(form = ~ Visit),
                  correlation = corAR1(form = ~ Visit | id),
                  method = "REML",
                  data = df)
summary(true.model)
plot(true.model)
```
\normalsize

## Hypothesis Testing

We perform a t-test for each of our regression parameters (coefficients) in our model
\footnotesize
```{r t_test}
summary(true.model)$tTable
```
\normalsize
COMMENT: Need to interpret the results

We test the coefficients in our model. We start with checking for different mean trajectories between the sexes. 
COMMENT: But why, since we already have our model. Some of these have different DFs and can't be compared, what does this mean? 

\footnotesize
```{r hypo_sex_test}
anova(true.model, Terms = c("Visit", "Visit:sex"))
```
\normalsize

































