---
title: "Group 7"
date: "2026-03-25"
output: 
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-2cm} 
fontsize: 9pt
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
load("Group_7.RData")
df <- DF

library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(nlme)
```

# Introduction

This study investigates the progression of motor symptoms in patients diagnosed with Parkinson's Disease (PD). The primary aim is to understand how smoking status influences the disease trajectory. In addition, the aim is to understand how this evolution is influenced by key patient characteristics such as sex, age at diagnosis.The study enrolled 500 patients who were diagnosed with PD between 1995 and 2011. Motor assessments were conducted at baseline and then annually for a total follow-up period of 7 years. The disease trajectory is assessed by motor symptoms measured using the Unified Parkinson's Disease Rating Scale (UPDRS) Part III: Motor Examination score. This score ranges from 0 to 108, where higher values indicate more severe motor impairment.

# Descriptive Analysis

```{r data_fix, include=FALSE}
df$sex_num <- ifelse(df$sex == "male", 0, 1) # For plotting purposes
df$id <- as.numeric(as.character(df$id))

df$age_groups <- cut(
  df$Age,
  breaks = c(40, 50, 60, 70, 80, 90),
  labels = c("40s", "50s", "60s", "70s", "80s"),
  include.lowest = TRUE,
  right = TRUE
)

df_wide <- df %>%
  pivot_wider(names_from = Visit,
              values_from = UPDRS,
              names_prefix = "UPDRS_")

df_wide <- df_wide[, c(1:9, 13, 10, 14, 12, 11)] # Rearrange the columns into the correct order
head(df_wide)
```

We have the following distributions in our data. The sexes of the participants are roughly equal, with just slightly more females than males. Age ranges from 41.1 to 88 and follows no clear distribution. The majority of our participants are non-smokers, with former smokers making up about half the number of non-smokers and current smokers just over a third the amount of non-smokers. Finally, at baseline we have $\sim500$ measurements, which drops to the range 306-339 for each subsequent visit. This means our data is unbalanced, which indicates that a continuous AR1, exponential serial, or Gaussian serial correlation structure will be needed.

```{r sum_stats, out.width="75%"}
plot1 <- ggplot(data = df, aes(x = sex_num)) +
  geom_histogram(stat = "bin", binwidth = 0.5) +
  stat_bin(binwidth = 0.5,
           aes(label=ifelse(after_stat(count)==0,"",after_stat(count))),
           geom="text", vjust=-.5) +
  ylim(0, 1525) + scale_x_continuous(breaks = c(0,1)) +
  labs(x = "Sex", y = "Count")

# plot2 <- ggplot(data = df, aes(x = Age)) +
#   geom_histogram(stat = "bin", binwidth = 0.5) +
#   ylim(0,85) +
#   labs(y = "Count")

plot2 <- ggplot(df, aes(x = age_groups)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ifelse(
    after_stat(count) == 0, "", after_stat(count)
  )), vjust = -0.5) +
  ylim(0, 725) +
  labs(x = "Age", y = "Count")

plot3 <- ggplot(df, aes(x = smoking_status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ifelse(
    after_stat(count) == 0, "", after_stat(count)
  )), vjust = -0.5) +
  ylim(0, 1600) +
  labs(x = "Smoking Status", y = "Count")

plot4 <- ggplot(df, aes(x = Visit)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ifelse(
    after_stat(count) == 0, "", after_stat(count)
  )), vjust = -0.5) +
  ylim(0, 550) +
  labs(y = "Count")

grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)
```

The mean UPDRS score per visit for each smoking subgroup (non-, former, and current smoker) appears non-linear. The standard error also increases for each visit.

```{r smoking_plot, out.width="75%"}
smokers.summary <- df %>% 
  group_by(smoking_status, Visit) %>% # Split the participants into their smoking status and evaluate the UPDRS by Visit
  summarise(.groups = "drop_last",
    mean_UPDRS = mean(UPDRS, na.rm=TRUE),
    sd_UPDRS = sd(UPDRS, na.rm=TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

split_by_smoking <- facet_wrap(~ smoking_status,  # Splits output into three graphs, one per smoking group
             labeller = labeller(smoking_status = c("0" = "Non-smoker", "1" = "Former smoker", "2" = "Current smoker")))

ggplot(smokers.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21, fill = "white", size = 2) +
  geom_errorbar(aes(ymin = mean_UPDRS - se_UPDRS,
                     ymax = mean_UPDRS + se_UPDRS), 
                width = 0.5, color = "blue") +
  # geom_smooth(method = "loess", color = "red", linewidth = 0.2) + # How to interpret result here?
  split_by_smoking +
  labs(x = "Visit", y = "UPDRS Score")
```

```{r sex_plot}
sex.summary <- df %>%
  group_by(sex_num, Visit) %>% # Split the participants by sex and evaluate the UPDRS by Visit
  summarise(
    .groups = "drop_last",
    mean_UPDRS = mean(UPDRS, na.rm = TRUE),
    sd_UPDRS = sd(UPDRS, na.rm = TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

sex.summary

split_by_sex <- facet_wrap( ~ sex_num, labeller = labeller(sex_num = c("0" = "Male", "1" = "Female")))

ggplot(sex.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21,
                                          fill = "white",
                                          size = 2) +
  geom_errorbar(
    aes(ymin = mean_UPDRS - se_UPDRS, ymax = mean_UPDRS + se_UPDRS),
    width = 0.5,
    color = "blue"
  ) +
  split_by_sex +
  labs(x = "Visit", y = "UPDRS Score")
```

```{r age_plot}
age.summary <- df %>%
  group_by(age_groups, Visit) %>% # Split the participants into their smoking status and evaluate the UPDRS by Visit
  summarise(
    .groups = "drop_last",
    mean_UPDRS = mean(UPDRS, na.rm = TRUE),
    sd_UPDRS = sd(UPDRS, na.rm = TRUE),
    n = sum(!is.na(UPDRS)),
    se_UPDRS = sd_UPDRS / sqrt(n)
  )

age.summary

# split_by_age <- facet_wrap(~ age_groups, labeller = labeller(sex_num = c("0" = "Male", "1" = "Female")))

ggplot(age.summary, aes(x = Visit, y = mean_UPDRS)) +
  geom_line(color = "black") + geom_point(shape = 21,
                                          fill = "white",
                                          size = 2) +
  geom_errorbar(
    aes(ymin = mean_UPDRS - se_UPDRS, ymax = mean_UPDRS + se_UPDRS),
    width = 0.5,
    color = "blue"
  ) +
  facet_wrap( ~ age_groups, nrow = 1) +
  labs(x = "Visit", y = "UPDRS Score")
```



This spread can be seen in the following spaghetti plots from a random sample of participants.

```{r spaghetti_plots, out.width="75%"}
# If I were better with R, I'd make this 3x times repeated object a function
# - One day we will get there :D
random_smokers <- df %>% distinct(smoking_status, id) %>%
  group_by(smoking_status) %>% slice_sample(n = 12)
df_sample_smokers <- df %>%
  semi_join(random_smokers, by = c("smoking_status", "id"))

random_sex <- df %>% distinct(sex_num, id) %>%
  group_by(sex_num) %>% slice_sample(n = 12)
df_sample_sex <- df %>%
  semi_join(random_sex, by = c("sex_num", "id"))

random_age <- df %>% distinct(age_groups, id) %>%
  group_by(age_groups) %>% slice_sample(n = 12)
df_sample_age <- df %>%
  semi_join(random_age, by = c("age_groups", "id"))

ggplot(df_sample_smokers,
       aes(
         x = Visit,
         y = UPDRS,
         group = id,
         colour = factor(id)
       )) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap( ~ smoking_status, nrow = 1) +
  split_by_smoking +
  labs(x = "Visit", y = "UPDRS Score")

ggplot(df_sample_sex, aes(
  x = Visit,
  y = UPDRS,
  group = id,
  colour = factor(id)
)) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap( ~ sex_num, nrow = 1) +
  split_by_sex +
  labs(x = "Visit", y = "UPDRS Score")

ggplot(df_sample_age, aes(
  x = Visit,
  y = UPDRS,
  group = id,
  colour = factor(id)
)) +
  geom_line(show.legend = FALSE, alpha = 0.75) +
  facet_wrap( ~ age_groups, nrow = 1) +
  labs(x = "Visit", y = "UPDRS Score")

#grid.arrange(plot5, plot6, ncol = 1)
```

# Statistical Analysis

The plot of squared OLS residuals against visit indicates that the variance is not constant, which means that we have heteroscedasticity. Thus, we get further evidence backing up the spaghetti plots that the progression of Parkinson's across the patients in this study is not uniform.

```{r ols_plot, out.width="75%"}
lin.reg <- lm(UPDRS ~ Visit, data = df)
sq.ols.residuals <- resid(lin.reg)^2

df2 <- df
df2$sq_ols_residuals <- sq.ols.residuals
df2$fitted <- fitted(lin.reg)

ggplot(df2, aes(x = Visit, y = sq_ols_residuals)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Visit", y = "Squared OLS Residuals")
```

From the correlation matrix of the available UPDRS Scores below, we get the expected result that UPDRS Scores are less correlated the further along the measurement was taken.
\footnotesize
```{r cor}
df_wide_UPDRS <- subset(df_wide, select = -c(1, 2, 3, 4, 5, 6))
as.tibble(cor(df_wide_UPDRS, use = "pairwise.complete.obs"))
```
\normalsize


SLIDE 268 ONWARDS FOR HERE. 


We obtain,through consideration of the research question and by comparing models with different interactions, our model as the following;

$$ y_{ij} = \big( \beta_0 + \text{b}_{i0} \big) + \big(\beta_1 + \text{b}_{i1}\big)\text{Visit}_{ij} + \beta_2\,\text{sex}_i + \beta_3\,\text{smoking}_i + \beta_4\,\text{age}_i + \beta_5\,\text{Visit}_{ij}\;\text{sex}_i + \beta_6\,\text{Visit}_{ij}\;\text{smoking}_i + \beta_7\,\text{sex}_{i}\;\text{smoking}_i + \beta_8\,\text{Visit}_{ij}\;\text{sex}_i\;\text{smoking}_i + \epsilon_{ij}$$

where $\beta$ are our model's regression parameters, $i = 1,2,\dots,2765$ denotes the number of participants, $j = 0,1,\dots,7$ denotes the number of visits, b$_{i0}$ denotes the random intercept of each patient, and b$_{i1}$ denotes the random slope of each patient's progression.

```{r model, echo=TRUE}
true.model <- lme(
  UPDRS ~ Visit * smoking_status * sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
summary(true.model)
```




# Appendix

## Model Selection

We started with a simple model with only an interaction between visit number and smoking and focus on whether our model needed random intercepts only, or both random intercepts and random slopes.

Discrete AR(1) (denoted by AR1) was chosen over continuous AR(1) (CAR1) as the visits themselves are equally spaced (which implies AR1), even thought the measurements are sometimes missing so the actual measurements aren't actually evenly spaced.

The package *nlme* was chosen to generate our model because *lme4* doesn't model heteroskedasticity and doesn't allow for a specified correlation structure, the latter not necessarily being a problem conceptually but ended up not working well owing to many subjects having lots of missing measurements.

First, we check the within-subject dependence of the basic model by comparing random intercepts, random slopes, and the AR1 correlation structure.

\footnotesize
```{r within_subject_dependence1}
m1 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = ~ 1 | id,
  # Visit | id
  weights = varExp(form = ~ Visit),
  # defines heteroskedasticity in Visit within the group.
  correlation = corAR1(form = ~ 1 |
                         id),
  # visits are equally spaced, but measurements are sometimes missing: AR1 or CAR1?
  method = "REML",
  data = df
)

m2 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  # Removes intercept-slope covariance (gets this model to work!)
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("AR1 & Random intercepts vs. AR1 & random intercepts + slopes")
anova(m1, m2)
print("Chosen: AR1 & random intercepts + slopes")
```
```{r within_subject_dependence2}
m1 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = ~ Visit | id,
  weights = varExp(form = ~ Visit),
  method = "REML",
  data = df
)

m2 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("Random intercepts + slopes vs. AR1 & random intercepts + slopes")
anova(m1, m2)
print("Chosen: AR1 & random intercepts + slopes")
```
\normalsize
As these models aren't nested, we consider the information criteria. The model with both random intercepts and slope (but with no intercept-slope covariance) and AR(1) has the lowest AIC and BIC, so we keep it.

We then checked for more interactions between the terms.
\footnotesize
```{r comp_int_nlme1}
m1 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)

m2 <- lme(
  UPDRS ~ Visit * smoking_status * sex * Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("Visit-smoking interaction only vs. All interactions")
anova(m1, m2)
print("Chosen: All interactions")
```
```{r comp_int_nlme2}
m1 <- lme(
  UPDRS ~ Visit * smoking_status * sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)

m2 <- lme(
  UPDRS ~ Visit * smoking_status * sex * Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("Age non-interacting vs. All interactions")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
```{r comp_int_nlme3}
# Remove interaction between age and everything else
m1 <- lme(
  UPDRS ~ Visit * smoking_status + sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
# All interactions
m2 <- lme(
  UPDRS ~ Visit * smoking_status * sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("Age-sex non-interacting vs. Age non-interacting")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
```{r comp_int_nlme4}
m1 <- lme(
  UPDRS ~ Visit * smoking_status + sex * Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)

m2 <- lme(
  UPDRS ~ Visit * smoking_status * sex + Age,
  random = list(id = pdDiag( ~ Visit)),
  weights = varExp(form = ~ Visit),
  correlation = corAR1(form = ~ Visit | id),
  method = "REML",
  data = df
)
print("Age-sex interacting with each other only vs. Age non-interacting")
anova(m1, m2)
print("Chosen: Age non-interacting")
```
\normalsize

This led to the selection of our true model.

\footnotesize
```{r true_model_nlme, echo=TRUE, out.width="75%"}
summary(true.model)
plot(true.model)
```
\normalsize

## Hypothesis Testing (Move out of appendix?)

We perform a t-test for each of our regression parameters (coefficients) in our model
\footnotesize
```{r t_test}
summary(true.model)$tTable
```
\normalsize
From the above results, it looks as if the only statistically significant terms of our model are the visit, sex, and the interaction between the two.

We test the coefficients in our model. We start with checking for different mean trajectories between the sexes, with the null hypothesis that an increase in visit does not have an effect, and that there is no difference in progression between males and females. Intuitively, we expect to reject the null hypothesis as we have seen in our exploration of the data at the start that there is a significant difference between the sexes. This is exactly the result we get below.

\footnotesize
```{r hypo_visit_sex_test}
anova(true.model, Terms = c("Visit", "Visit:sex"))
```
\normalsize

We now check the null hypothesis that smoking does not have a statistically significant effect on the outcome. We recall from our exploration that all particpiants in each of the statuses had a positive non-linear trajectory when mapping visit versus UPDRS score, though with former smokers having the steepest curve and non-smokers have the least. Our intuition is that whether or not one smokes, smokes, or once did so won't have a large impact on the UPDRS score, and this is exactly the outcome we get below. 

```{r hypo_smoking_test}
anova(true.model, Terms = "smoking_status")
```
Note, however, that this term stays in our model as it was specifically requested for analysis in the research question.

Finally, we check the hypothesis that age does not have a significance on the outcome of the UPDRS score of our participants. From our exploration of the data, we saw that there was some variability between the age groups, though this was not in the expected way (namely, that the UPDRS scores were higher on average for older people) as we got that those in their 60s had a lesser progression than those in their 50s. This doesn't seem to correspond to a lack or excess of data. Thus, checking our hypothesis, we get the result that age has a weak effect on the UPDRS score.

```{r hypo_age_test}
anova(true.model, Terms = "Age")
```




























